---
title: "More_example_code"
author: "Amelia Mesich"
date: "2025-09-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up

Time to test your hypothesis! First we'll load all the libraries required and the data with for all the species.

```{r packages}
# Load the libraries required for analysis and visualization
library(ggplot2)
library(lubridate)
library(dplyr)
#install.packages("ggpubr")
library(ggpubr)

# Load the data patterns for all of the species 
# This will appear in your environment as a dataframe called "all_species_sleep_data"
load("Tutorial4.RData")
View(all_species_sleep_data)
```

Filter the data for the species you've chosen to compare! Replace each text with the 6 letter species code corresponding to your species. Here, the \| symbol is an "or" operator, which means we want to filter the data where the species name is X or Y. If you've chosen more species to look at, you can just add another or (\|) to include it! In this example instead of two species we'll look at four.

If you're looking at total sleep or sleep bouts you need to filter by state == sleep. Unless you're also looking at wake metrics for your hypothesis you'll need to filter for just the sleep observations.

```{r filter data}
my_species <- filter(all_species_sleep_data, species == "Neomul" | species == "Cypcol" | species == "Boumic" | species == "Lepatt") %>% filter(state == "sleep")

```

## Testing your hypothesis

Here, you need to clean up your data to focus only on the information needed for your hypothesis, like we did in Tutorial 3. You can use the group_by and summarise functions to create a new dataframe called sleep_analysis. Helpful functions to summarise by include mean(variable) and sum(variable).

If you're looking at total sleep you'll need to sum the total hours. If you're interested in the total number of bouts in a phase you can also use sum. If you're looking at any of the other metrics (number of bouts, bout length, longest bout, y position), you'll need to take the mean.

```{r analysis}
#colnames shows what each of the columns are called in the dataframe
colnames(all_species_sleep_data)

sleep_analysis <- my_species %>% 
  group_by(species, sample_id, sex, diet) %>%
  summarise(mean_bouts = mean(n_bouts), total_hours_sum = sum(total_hour)) 

head(sleep_analysis)
```

## Visualizing your data

Once you have a dataframe containing only the variables of interest, you can use ggplot to make all kinds of graphs. The outline code here is for a boxplot (which will be useful), but we encourage you to also explore other plots to include in your graphical abstract! You can re-use code from Tutorial 2, where we make line graphs, histograms and scatterplots. There is also useful code there to personalize the graphs, including annotations, changing colours, etc.

Make sure when you're ploting your graph you use the same variable names as are in your dataframe called sleep_analysis. You can check this with head(sleep_analysis) or colnames(sleep_analysis).

## Comparison of two groups

If you're only comparing two groups you don't need to include the comparisons argument in stat_compare_means since R will automatically perform a t test on the two groups present in the data.

```{r plotting}
ggplot(sleep_analysis, aes(x = diet, y = total_hours_sum)) +
  geom_boxplot() +
  geom_point(position = position_dodge(0.75)) +
  stat_compare_means(method = "t.test", paired = FALSE, label="p") +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y")

```

Here we can see there is no statistically significant difference in total sleep between piscivores and zooplanktivores.

## Comparison of multiple groups

Next we'll show an example of how to add the statistical analysis for multiple species. You'll need to create a list with each of the comparisons you want to include when testing for differences in group means.

Also, since we are separating the boxplots by colour we need to separate out the points in geom_point() to match using position_dodge(0.75).

```{r plotting}
comparison_list <- list(c("Boumic", "Cypcol"), c("Boumic", "Lepatt"), c("Boumic", "Neomul"), c("Cypcol", "Lepatt"), c("Cypcol", "Neomul"), c("Lepatt", "Neomul"))

ggplot(sleep_analysis, aes(x = species, y = total_hours_sum, colour = diet)) +
  geom_boxplot() +
  geom_point(position = position_dodge(0.75)) +
  stat_compare_means(method = "t.test", paired = FALSE, label="p", comparison = comparison_list) +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y")

```

## Facet_wrapping

If you want to compare groups by an additional variable you can separate the data into separate plots using facet_wrap(\~additional_variable_name). This is an example separating the individuals by sex.

```{r plotting}

ggplot(sleep_analysis, aes(x = species, y = total_hours_sum, colour = diet)) +
  geom_boxplot() +
  geom_point(position = position_dodge(0.75)) +
  stat_compare_means(method = "t.test", paired = FALSE, label="p", comparisons = comparison_list) +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y") + facet_wrap(~sex, scales = "free")

```

Another useful way to look at differences within a species is to colour or fill by a variable of interest

```{r plotting2}

ggplot(sleep_analysis, aes(x = species, y = total_hours_sum, fill = sex)) +
  geom_boxplot() +
  geom_point(position = position_dodge(0.75)) +
  stat_compare_means(method = "t.test", paired = FALSE, label="p") +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y") 

```
## Plotting differences in phase

```{r analysis}
#colnames shows what each of the columns are called in the dataframe
colnames(all_species_sleep_data)

sleep_analysis <- my_species %>% 
  group_by(species, sample_id, phase) %>%
  summarise(mean_bouts = mean(n_bouts), total_hours_sum = sum(total_hour)) 

head(sleep_analysis)
```

We can look at the total sleep in each phase for Boumic. Since we're comparing total sleep in each phase, our comparison list will need to include each of the phases instead of species names. You could then repeat this filtering for a different species to see if another species has a different phase preference.

```{r plotting}
boumic_sleep_analysis <- filter(sleep_analysis, species == "Boumic")

comparison_list <- list(c("dawn", "day"), c("dawn", "dusk"), c("dawn", "night"), c("day", "dusk"), c("day", "night"), c("dusk", "night"))

ggplot(boumic_sleep_analysis, aes(x = phase, y = total_hours_sum, colour = phase)) +
  geom_boxplot() +
  geom_point(position = position_dodge(0.75)) +
  stat_compare_means(method = "t.test", paired = FALSE, label="p", comparison = comparison_list) +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y")

```

## Removing outliers

If you have some outliers that are heavily skewing your data you can remove them by setting outlier.shape = NA in the geom_boxplot() layer. You will also need to remove the geom_point() layer since this layer will add the outliers back when mapping out all the points. Additionally you can set the upper and lower limits of the y axis using coord_cartesian(ylim = c(lower_limit, upper_limit)) to manually fix the scaling. Only do this if an outlier is heavily skewing your data and making it difficult to visualize.

```{r plotting}
my_species <- filter(all_species_sleep_data, species == "Ophboo" | species == "Lepatt")

sleep_analysis <- my_species %>% 
  group_by(species, sample_id, sex) %>%
  summarise(mean_longest_bout = mean(longest_bout)) 

#colouring the outliers red so you can see them
ggplot(sleep_analysis, aes(x = species, y = mean_longest_bout)) +
  geom_boxplot(outlier.colour = "red") +
  stat_compare_means(method = "t.test", paired = FALSE, label="p") +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y")

#removing the outliers
ggplot(sleep_analysis, aes(x = species, y = mean_longest_bout)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(method = "t.test", paired = FALSE, label="p") +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y") +
  coord_cartesian(ylim =  c(0, 15000))

```
