---
title: "Tutorial_1.R"
author: "Amelia Mesich"
date: "2025-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome to R!

R is the programming language we will use for the next few weeks of tutorials.

R allows you to set the value of variables and perform functions with them

```{r introduction}
2+2
a = 2 + 3
x = 4
y = 7
x + y
```

## Classes of objects

There are various types of data recognized by R. These including numerics, characters, lists and more.

```{r object classes}
class(2)
class("hello")
class(list("I", "love", "sleep"))
class(list(2,5,3,9,1))

```

We will mostly be working with dataframes. These consist of rows and columns of data.

```{R dataframe}
data.frame(numbers = c("A", "B", "C", "D"), integers = c(1,2,3,4))

```

##R has lots of other built in functions

```{r functions}
print("Hello world")
max(2,5,3,9,1)
min(2,5,3,9,1)

```

##R can be used to analyze and visualize data

```{r plotting, echo=TRUE}
plot(x = c(1,2,3,4), y = c(5, 10, 2, 5), type = "b")
```

#See you next week!

```{r fish plot, echo=TRUE}
example_x <- c(1,1,2,4,5,6,7,8,8,7,6,5.5,5,4,4,4,2,1)
example_y <- c(4,12,9,12,12,12,12,10,7,5,5,5,1,5,5,5,7,4)
plot(x = example_x, y = example_y, type = "l", col = "blue", lwd = 4)
symbols(6.5,9, circles = 0.1, add = TRUE, bg = "red", lwd = 4, inches = 0.3)
symbols(6.5,9, circles = 0.1, add = TRUE, bg = "black", lwd = 4, inches = 0.1)
```
