---
title: "Tutorial_4_starter_code"
author: "Ayasha Abdalla-Wyse"
date: "2025-09-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up

Time to test your hypothesis! First we'll load all the libraries required and the data with for all the species.

```{r packages}
# Load the libraries required for analysis and visualization
library(ggplot2)
library(lubridate)
library(dplyr)

# Load the data patterns for all of the species 
# This will appear in your environment as a dataframe called "all_species_sleep_data"
load("Tutorial4.RData")
```

Filter the data for the species you've chosen to compare! Replace each text with the 6 letter species code corresponding to your species. Here, the \| symbol is an "or" operator, which means we want to filter the data where the species name is X or Y. If you've chosen more species to look at, you can just add another or (\|) to include it!

```{r filter data}
my_species <- filter(all_species_sleep_data, species == "______" | species == "______")

```

## Testing your hypothesis

Here, you need to clean up your data to focus only on the information needed for your hypothesis, like we did in Tutorial 3. You can use the group_by and summarise functions to create a new dataframe called sleep_analysis. Helpful functions to summarise by include mean(variable) and sum(variable).

```{r analysis}
sleep_analysis <- my_species %>% 
  group_by(species, sample_id, some_variable, ...) %>%
  summarise(new_column = ____) 

head(sleep_analysis)
```

## Visualizing your data

Once you have a dataframe containing only the variables of interest, you can use ggplot to make all kinds of graphs. The outline code here is for a boxplot (which will be useful), but we encourage you to also explore other plots to include in your graphical abstract! You can re-use code from Tutorial 2, where we make line graphs, histograms and scatterplots. There is also useful code there to personalize the graphs, including annotations, changing colours, etc.

```{r plotting}
ggplot(sleep_analysis, aes(x = ____, y = ____)) +
  geom_boxplot() +
  geom_point() +
  stat_compare_means(method = "t.test", paired = FALSE, label="p") +
  ggtitle("TITLE") +
  labs(x = "Label x", y = "Label y")
```
